---
title: "Hot-spots and Cold-spots"
subtitle: "Preliminary Analysis"
author: "Timothy Fraser, Aldrich Resilience Lab, UNDP AccLab Paraguay"
date: "July 7, 2020"
output: html_notebook
---

This document generates maps using spatial interpolation to describe requests for aid and provision of aid in Paraguay during COVID-19.

# 0. Packages and Data

```{r, message = FALSE, warning = FALSE}
library(tidyverse) # for data manipulation
# Geospatial packages
library(sf)
library(rgdal)
library(rgeos)
library("gstat")
library("sp")
library("maptools")
# Color palletes
library(viridis)
library(wesanderson)

# South American Albers Equal Area Conic projection
# Source: https://spatialreference.org/ref/esri/south-america-albers-equal-area-conic/
saea <- "+proj=aea +lat_1=-5 +lat_2=-42 +lat_0=-32 +lon_0=-60 +x_0=0 +y_0=0 +ellps=aust_SA +units=m +no_defs"

# Create directory for GIS shape data
dir.create("shapes")

# Global Administrative Areas dataset from UC Davis
# https://gadm.org/download_country_v3.html
#download.file(url = "https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_PRY_2_sp.rds",
#              destfile = "shapes/paraguay_bounds.rds")
```


# 1. Generate key polygons and points

## 1.1 Import Paraguay Polygons

```{r}
# Generate district level shapefiles for Paraguay
geo <- read_rds("shapes/paraguay_bounds.rds") %>% 
  st_as_sf() %>%
  st_transform(crs = 4326) %>%
  st_transform(crs = saea) %>%
  #st_set_crs(saea) %>%
  dplyr::select(dept = NAME_1, district = NAME_2) %>%
  group_by(dept, district) %>%
  summarize(geometry = st_union(geometry)) %>%
  ungroup() %>%
  mutate(id = 1:n())
```

## 1.2 Geocode requests for aid and organziation that provide aid.

```{r}
# Let's geocode all instances of requests from Ayudapy.org
read_rds("ayudapy_requests.rds") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = saea) %>%
  # now join in geographic identifier
  st_join(geo) %>%
  mutate(id = as.numeric(id)) %>%
  # and filter to only cases located within Paraguay
  filter(!is.na(district)) %>%
  # Now specify an outcome value (1 = presence)
  mutate(outcome = 1) %>%
  saveRDS("shapes/ayudapy_points.rds")

# Now let's geocode organizations from ElMapaSocial
read_csv("/cloud/project/shapes/ong.csv") %>%
  filter(!is.na(longitud) & !is.na(latitud)) %>%
  st_as_sf(coords = c("longitud", "latitud"), crs = 4326) %>%
  st_transform(crs = saea) %>%
  mutate(id = as.numeric(id)) %>%
  # and filter to only cases located within Paraguay
  st_join(geo %>% select(-id)) %>%
  filter(!is.na(district)) %>%
  mutate(outcome = 1) %>%
  saveRDS("shapes/elmapasocial_points.rds")

# Now let's geocode organizations from Ushahide
read_csv("/cloud/project/shapes/ushahide.csv") %>% 
  dplyr::select(
    id = `Post ID`,
    survey = Survey,
    longitude = `Ubicación aproximada (si aplica).lon`,
    latitude = `Ubicación aproximada (si aplica).lat`) %>%
  mutate(id = as.numeric(id)) %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = saea) %>%
  st_join(geo %>% select(-id)) %>%
  filter(!is.na(district)) %>%
  mutate(outcome = 1) %>%
  saveRDS("shapes/ushahide_points.rds")


# Combine points
bind_rows(
  # Ushahide
  read_rds("shapes/ushahide_points.rds") %>% 
    mutate(source = "Ushahide",
           type = "Volunteer Org") %>%
    dplyr::select(id, outcome, source, type, geometry) %>%
    mutate(id = as.numeric(id)),
  # El Mapa Social
  read_rds("shapes/elmapasocial_points.rds") %>% 
    dplyr::select(id, is_aprobado, is_donacion, is_voluntariado, is_adoptar) %>%
    pivot_longer(
      cols = c(is_aprobado, is_donacion, is_voluntariado, is_adoptar),
      names_to = "type",
      values_to = "outcome") %>%
    mutate(type = type %>% dplyr::recode(
      "is_aprobado" = "Other Org",
      "is_adoptar" = "Other Org",
      "is_donacion" = "Charitable Org",
      "is_voluntariado" = "Volunteer Org")) %>%
    distinct() %>%
    mutate(source = "El Mapa Social") %>%
    filter(outcome == 1) %>%
    mutate(id = as.numeric(id)),
  # Aduyapy Points
  read_rds("shapes/ayudapy_points.rds") %>% 
    mutate(source = "Aduyapy",
           type = "Requests for Assistance",
           outcome = 1) %>%
    dplyr::select(id = order, source, type, outcome, geometry) %>%
    mutate(id = as.numeric(id))
) %>%
  mutate(id = 1:n()) %>%
  mutate(type = type %>% factor(levels = c(
    "Requests for Assistance", "Volunteer Org", "Charitable Org", "Other Org"))) %>%
  saveRDS("shapes/points.rds")

#read_rds("shapes/points.rds") %>% head()
```


# 2. Mapping Requests for Aid and Provison of Aid

```{r}
# Basic visual
ggplot() +
  # map polygon boundaries
  geom_sf(data = geo, mapping = aes(geometry = geometry),
          fill = "darkgrey", color = "grey") +
  # map points, specifying type of point by color
  geom_sf(data = read_rds("shapes/points.rds"),
          mapping = aes(geometry = geometry,
                        color = type),
          alpha = 0.5) +
    theme_void(base_size = 14) +
  # use this eye-catching pallete
  scale_color_manual(values = wesanderson::wes_palette(
    name = "Zissou1", 4, type = "continuous")) +
  labs(title = "Requests and Offers of Aid",
       color = "Type") +
  theme(plot.title = element_text(hjust = 0.5))
```

## 2.1 Fishnet

Next, let's create a fishnet - a grid that tallies the number of requests per fixed geographic distance.

```{r}
fishnet <- st_make_grid(
  st_buffer(
    geo, 1),
  # 10 square kilometer grid cells
  # (unit is meters)
  cellsize = 10000,
  what = "polygons") %>%
  st_as_sf() %>%
  mutate(id = 1:n()) %>%
  dplyr::rename(geometry = x) %>%
  # Now join in the points, and tally how many occur per grid-cell
  st_join(read_rds("shapes/points.rds") %>% 
            dplyr::select(outcome, type, geometry)) %>%
  group_by(id, type) %>%
  summarize(outcome = sum(outcome, na.rm = TRUE)) %>%
  ungroup()
```


## 2.2 Visualize Fishnet


```{r}
# Get department and country polygon
dept <- geo %>%
  group_by(dept) %>%
  summarize(geometry = st_union(geometry))

country <- geo %>%
  summarize(geometry = st_union(geometry))
```

```{r}
ggplot() +
  geom_sf(data = fishnet, fill = "white", size = 0.1,
          alpha = 0.5, color = "black") +
  geom_sf(data = read_rds("shapes/points.rds"),
          mapping = aes(geometry = geometry,
                        color = type),
          alpha = 0.5) +
    theme_void(base_size = 16) +
  geom_sf(data = country, color = "black", fill = NA) +
  scale_color_manual(
    values = wesanderson::wes_palette(name = "Zissou1", 4, type = "continuous")) +
  scale_fill_gradient2(low = "black", high = "grey") +
 # scale_fill_grey() +
  theme_void() +
  guides(fill = guide_colourbar(barwidth = 0.5, barheight = 5)) +
  labs(color = "Type")
```

## 2.3 Spatial Interpolation

```{r, message = FALSE, warning = FALSE}

# Make a models for each type
requests <- gstat(id = "id", formula = outcome ~ 1, 
            data = fishnet %>% 
             filter(type == "Requests for Assistance") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

volunteer <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Charitable Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

charitable <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Volunteer Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

other <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Other Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

# Make a grid of polygons to predict
nodes <- st_make_grid(
  st_buffer(
  geo, 1),
  cellsize = 10000,
  what = "polygons")

system.time(
  # Generate predictions
  bind_rows(
    predict(requests, nodes) %>%
      dplyr::rename(predicted = id.pred) %>%
      mutate(type = "Requests for Assistance"),
    
    predict(volunteer, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Volunteer Org"),
    
    predict(charitable, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Charitable Org"),
    
    predict(other, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Other Org")) %>%
    saveRDS("shapes/predictions.rds"))

# Remove excess data upon completion
remove(nodes, other, volunteer, charitable, requests)
```


```{r}

dept <- geo %>%
  group_by(dept) %>%
  summarize(geometry = st_union(geometry))

country <- geo %>%
  summarize(geometry = st_union(geometry))
```

## 2.4 Compare predictions vs. actual

```{r}
ggplot() +
  # Map predictions in 10-square-kilometer cells
  geom_sf(data = read_rds("shapes/predictions.rds"), 
          mapping = aes(fill = log(predicted)), 
          size = 0.5, color = NA) +
  # Map points by color to compare predictions vs. actual
  geom_sf(data = read_rds("shapes/points.rds"), 
          mapping = aes(color = type),
          alpha = 0.5) +
   scale_color_manual(
    values = wesanderson::wes_palette(name = "Zissou1", 4, type = "continuous")) +
  scale_fill_gradient(low = "black", high = "white") +
  # Add border data
  geom_sf(data = dept, fill = NA, color = "darkgrey") +
  geom_sf(data = country, fill = NA, color = "black") +
  theme_void() +
  facet_wrap(~type) +
  guides(fill = guide_colorbar(barwidth = 0.5, barheight = 5)) +
  labs(title = "Predicted Requests for Aid vs. Social Orgs",
       color = "Type",
       fill = "Predicted Number (log)")
```



```{r}
# Also, here's a histogram if you want it
read_rds("shapes/predictions.rds") %>%
  ggplot(mapping = aes(x = predicted, fill = type)) +
  geom_histogram() +
  facet_wrap(~type, scales = "free") +
  scale_fill_manual(values = wes_palette(name = "Zissou1", n = 4, type = "continuous"))
```

# 2.5 Get district level average predictions

```{r}
# Join predictions into shapes,
# and then summarize them by district
geo %>%
  st_join(read_rds("shapes/predictions.rds")) %>% 
  group_by(district, dept) %>%
  summarize(avg_predicted = mean(predicted, na.rm = TRUE)) %>%
  ungroup() %>%
  saveRDS("shapes/avg_predictions.rds")

# Visualize average predictions per district
ggplot() +
  geom_sf(data = read_rds("shapes/avg_predictions.rds"), 
          mapping = aes(fill = log(avg_predicted)), size = 0.5, color = NA) +
  geom_sf(data = read_rds("shapes/points.rds"), 
          mapping = aes(color = type),
          alpha = 0.5) +
   scale_color_manual(
    values = wesanderson::wes_palette(name = "Zissou1", 4, type = "continuous")) +
  scale_fill_gradient(low = "white", high = "black") +
  # Add border data
  geom_sf(data = dept, fill = NA, color = "darkgrey") +
  geom_sf(data = country, fill = NA, color = "black") +
  theme_void() +
  facet_wrap(~type) +
  labs(title = "Predicted Requests for Aid vs. Social Orgs",
       color = "Type",
       fill = "Predicted Number (log)")
```




# 3. Compare High Vulnerability vs. High Social Resource areas

```{r}
ggplot() +
  geom_sf(data = read_rds("shapes/avg_predictions.rds"), 
          mapping = aes(fill = log(avg_predicted)), size = 0.5, color = NA) +
  geom_sf(data = read_rds("shapes/points.rds"), 
          mapping = aes(color = type),
          alpha = 0.5) +
   scale_color_manual(
    values = wesanderson::wes_palette(name = "Zissou1", 4, type = "continuous")) +
  scale_fill_gradient(low = "white", high = "black") +
  # Add border data
  geom_sf(data = dept, fill = NA, color = "darkgrey") +
  geom_sf(data = country, fill = NA, color = "black") +
  theme_void() +
  facet_wrap(~type) +
  labs(title = "Predicted Requests for Aid vs. Social Orgs",
       color = "Type",
       fill = "Predicted Number (log)")
```

## 3.1 Generate requests and provision indices from 0 to 1

```{r}
request <-  read_rds("shapes/predictions.rds") %>%
  group_by(type) %>%
  mutate(id = 1:n()) %>%  ungroup() %>%
    filter(type == "Requests for Assistance") %>% 
    dplyr::select(id, requests = predicted, geometry)
  
aid <-  read_rds("shapes/predictions.rds") %>%
    group_by(type) %>%
  mutate(id = 1:n()) %>%  ungroup() %>%
    filter(type != "Requests for Assistance") %>%
    group_by(id) %>%
    summarize(predicted = sum(predicted,na.rm = TRUE)) %>%
    dplyr::select(id, social_orgs = predicted, geometry)


dat <- request %>%
  left_join(by = "id", 
            y = aid %>% as.data.frame() %>% 
              select(id, social_orgs, -geometry)) %>%
  # Convert into index
  mutate(requests = log(requests) %>% 
           scales::rescale(to = c(0,1)) %>% as.numeric,
         social_orgs = log(social_orgs) %>% 
           scales::rescale(to = c(0,1)) %>% as.numeric) %>%
  # Find difference in indices
  mutate(diff = requests - social_orgs)

# If there are more requests than social orgs, 
# relative to the distribution of both in the country,
# this means this is a place we would expect is getting underserved

# If the difference is POSITIVE (BLUE),
# that means there are more requests than social orgs.
# If the difference is NEGATIVE (RED),
# that means that there are more social orgs than requests

# Compare difference in requests vs. aid to see hotspots and coldspots
ggplot() +
  geom_sf(data = dat, mapping = aes(fill = diff), color = NA) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue") +
  theme_void(base_size = 14) +
  geom_sf(data = dept, fill = NA, color = "grey") +
  geom_sf(data = country, fill = NA, color = "black") +
  guides(fill = guide_colourbar(barwidth = 0.5, barheight = 10)) +
  labs(fill = "Difference in\nRequests\nvs.\nSocial Orgs",
       caption = "Requests for Aid and Social Orgs were tallied, \n predicted for every 10 square kilometers,\nlog-transformed, and rescaled from 0 to 1.\nSources: Ayudapy.org, ElMapaSocial.org, Py.Mapa.FrenalaCurva.net") +
  theme(plot.caption = element_text(hjust = 0.5))
```

# 4. City Specific Mapping

```{r}
capital <- geo %>%
  filter(dept == "Asunción")

# Generate a smaller grid for focusing just on the capital
fishnet <- st_make_grid(
  st_buffer(
    capital, 1),
  # 500 meter grid cells
  cellsize = 500,
  what = "polygons") %>%
  st_as_sf() %>%
  mutate(id = 1:n()) %>%
  dplyr::rename(geometry = x) %>%
  st_join(read_rds("shapes/points.rds") %>% 
            dplyr::select(outcome, type, geometry)) %>%
  group_by(id, type) %>%
  summarize(outcome = sum(outcome, na.rm = TRUE)) %>%
  ungroup()

# Make a models for each type
requests <- gstat(id = "id", formula = outcome ~ 1, 
            data = fishnet %>% 
             filter(type == "Requests for Assistance") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

volunteer <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Charitable Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

charitable <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Volunteer Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

other <- gstat(id = "id", formula = outcome ~ 1, 
           data = fishnet %>% 
             filter(type == "Other Org") %>%
             na.omit() %>%
             as("Spatial"),
            set = list(idp = 2))

# Make a grid of polygons to predict
nodes <- st_make_grid(
  st_buffer(
  capital, 1),
  cellsize = 500,
  what = "polygons")

system.time(
  # Generate predictions
  bind_rows(
    predict(requests, nodes) %>%
      dplyr::rename(predicted = id.pred) %>%
      mutate(type = "Requests for Assistance"),
    
    predict(volunteer, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Volunteer Org"),
    
    predict(charitable, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Charitable Org"),
    
    predict(other, nodes) %>%
      dplyr::select(predicted = id.pred) %>%
      mutate(type = "Other Org")) %>%
    saveRDS("shapes/predictions_capital.rds"))

request <-  read_rds("shapes/predictions_capital.rds") %>%
  group_by(type) %>%
  mutate(id = 1:n()) %>%  ungroup() %>%
    filter(type == "Requests for Assistance") %>% 
    dplyr::select(id, requests = predicted, geometry)
  
aid <-  read_rds("shapes/predictions_capital.rds") %>%
    group_by(type) %>%
  mutate(id = 1:n()) %>%  ungroup() %>%
    filter(type != "Requests for Assistance") %>%
    group_by(id) %>%
    summarize(predicted = sum(predicted,na.rm = TRUE)) %>%
    dplyr::select(id, social_orgs = predicted, geometry)


dat <- request %>%
  left_join(by = "id", 
            y = aid %>% as.data.frame() %>% 
              select(id, social_orgs, -geometry)) %>%
  # Convert into index
  mutate(requests = log(requests) %>% 
           scales::rescale(to = c(0,1)) %>% as.numeric,
         social_orgs = log(social_orgs) %>% 
           scales::rescale(to = c(0,1)) %>% as.numeric) %>%
  # Find difference in indices
  mutate(diff = requests - social_orgs)

remove(request, aid)


# Gather Paraguay streets for a better visual

#download.file(url = "https://github.com/justinelliotmeyers/official_paraguay_administrative_boundary_shapefile/raw/master/paraguay_2012_vias_completo.zip",
#              destfile = "shapes/capital_streets.zip")
# unzip("shapes/capital_streets.zip")

read_sf("shapes/capital_streets/paraguay_2012_vias_completo.shp") %>% 
  st_as_sf() %>%
  dplyr::select(cartodb_id, geometry) %>%
  st_transform(crs = 4326) %>%
  st_transform(crs = saea) %>%
  st_join(capital) %>%
  filter(!is.na(district)) %>%
  saveRDS("shapes/streets.rds")


# Visualize difference in aid requests vs. social org density
ggplot() +
  # visualize predictions per 500 meter block
  geom_sf(dat, mapping = aes(geometry= geometry, fill = diff), color = NA) +
  # Overlay Asuncion streets
  geom_sf(read_rds("shapes/streets.rds"), mapping = aes(geometry = geometry),
          color = "black", size = 0.1) +
  # Overlay border
  geom_sf(capital, mapping = aes(geometry = geometry), 
          color = "black", fill = NA) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue") +
  theme_void(base_size = 14) +
  guides(fill = guide_colorbar(barwidth = 0.5, barheight = 10)) +
  labs(fill = "Difference\nin\nRequests\nvs.\nSocial Orgs")
```





